\documentclass[a4paper, 11pt, normalem]{report}

\usepackage{../../../LaTeX-Templates/Notes}
\usepackage{subfiles}
\usepackage{tensor}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.pathmorphing}
\tikzset{snake it/.style={decorate,decoration=snake}}

\newcommand\xo{\hat{X}_1}
\newcommand\xt{\hat{X}_2}
\newcommand\hsig{\hat{\sigma}}
\newcommand\hd{\hat{\unl{d}}}
\newcommand\hrho{\hat{\rho}}
\newcommand\vk{\unl{k}}

\title{General Relativity \vspace{-20pt}}
\author{Richard Bower}
\date{\vspace{-15pt}Epiphany Term 2020}
\rhead{\hyperlink{page.1}{Contents}}

\begin{document}

\maketitle
\tableofcontents
\chapter{}
Just intro stuff

\chapter{Introduction to Tensors}
\begin{itemize}
    \item Notation
    \item Coordinate transforms
    \item Contravariant tensors
    \item Covariant tensors
\end{itemize}

\section{Intro to Tensor Notation}
Consider the cartesian definition for $\vr$:
\begin{align}
    \vr = x\vec{i} + y\vec{j} + \vec{z}.
\end{align}
We have the basis vector $\{\vec{i},\vec{j},\vec{k}\}$ and coordinate values $\{x,y,z\}$.
We can write this in a different form as
\begin{align}
    \vr = x^1\ve_1 + x^2\ve_2 + x^3\ve_3.
\end{align}
Note: $x^2 \neq x*x$.
The 2 is an index, not a power. 
If we want to square something, we will write $(x^1)^2 = x^1x^1$.
We can rewrite the above again as
\begin{align}
    \vr = \sum_{i=1}^3 x^i\ve_i.
\end{align}
We can then simplify this further using the \textbf{Einstein summation convention}:
\begin{align}
    \vr = x^i\ve_i,
\end{align}
i.e. whenever there is a repeated index, we sum over them. 
Different letters will imply different things:
\begin{itemize}
    \item Roman letters $i,j,\dots$ - summing over 3D space
    \item Roman letters $a,b,c,\dots$ - summing over ND space
    \item Roman letters $A,B,\dots$ - summing over 2D space
    \item Greek letters $\alpha,\beta,\mu,\nu,\dots$ - summing over 4D space-time $\{x^0,x^1,x^2,x^3\}$, starting from 0 as time is different slightly, so $\{ct,x^i\}$
\end{itemize}

\section{Coordinate Transformation}
You may be used to 
\begin{align}
    x' = \gamma\left(x-\frac{vct}{c}\right),
\end{align}
where the extra $c$ factor to make time space-like. 
This notation can get confusing so instead we use:
\begin{align}
    x^{\bar{1}} = \gamma\left(x^1-\frac{v}{c}x^0\right),
\end{align}
where the 'bar' indicates new coordinate system. 

For a minute vector difference between points P and Q $d\vr$ in two coordinate systems, we can define $\ve_a$:
\begin{align}
    \vr(P) &= \ve_ax^a & \vr(P) &= \ve_{\bar{b}}x^{\bar{b}} \\
    d\vr &= dx^a\ve_a  \\
    \frac{\p\vr}{\p x^a} &= \ve_a & \frac{\p\vr}{\p x^{\bar{b}}} &= \ve_{\bar{b}}
\end{align}
But what is the relationship between these two coordinate systems?
Start with $x^{\bar{b}}=x^{\bar{b}}(x^a)$, and consider a general function 
\begin{align}
    f &= f(x^1,x^2,x^3) \\
    \Delta f &= \frac{\p f}{\p x^1}\Delta x' + \frac{\p f}{\p x^2}\Delta x^2 + \frac{\p f}{\p x^2}\Delta x^3 = \frac{\p f}{\p x^a}\Delta x^a
\end{align}
How do we get a small change in $x^{\bar{b}}$?
\begin{align}
    \Delta x^{\bar{b}} &= \frac{\p x^{\bar{b}}}{\p x^{a}}\Delta x^{a} \\
    dx^{\bar{b}} &= \frac{\p x^{\bar{b}}}{\p x^a}dx^a \\
    dx^{\bar{a}} &= \frac{\p x^{\bar{a}}}{\p x^b}dx^b
\end{align}
Notice how we can simply just switch round the indices - \textbf{these are all dummy variables and as long as the index notation is consistent, it is completely arbitrary which letter is used,} i.e. the letters themselves mean nothing.

\section{Tensors}
Any quantity which transforms as 
\begin{align}
    A^{\bar{b}} &= \frac{\p x^{\bar{b}}}{\p x^a}A^a
\end{align} 
is a Rank (1,0) or order 1 contravariant tensor.
What about $\ve_a$?
\begin{align}
    \vr &= x^a\ve_a = x^{\bar{b}}\ve_{\bar{b}} \\
    \ve_{\bar{b}} &= \frac{\p\vr}{\p x^{\bar{b}}} = \frac{\p\vr}{\p x^a} \frac{\p x^a}{\p x^{\bar{B}}} = \frac{\p x^a}{\p x^{\bar{b}}}\ve_a
\end{align}
So now we have reversed the position of the indices in Eq (2.15).

How do we define scalars? 
\begin{align}
    \del\phi &= \frac{\p\phi}{\p x^i}\ve_i \\
    \frac{\p\phi}{\p x^{\bar{j}}} &= \frac{\p x^i}{\p x^{\bar{j}}}\frac{\p\phi}{\p x^i}
\end{align}
In general, we have
\begin{align}
    A_{\bar{j}} &= \frac{\p x^i}{\p x^{\bar{j}}}A_i,
\end{align}
which we call a Rank (0,1) or order 1 covariant tensor.

\chapter{}

\section{Higher order tensors}
Consider
\begin{align}
    T^{ab} &= A^aB^b, \\
    T^{\bar{c}\bar{d}} &= A^{\bar{c}}B^{\bar{d}} = \left(\frac{\p x^{\bar{c}}}{\p x^a}A^a\right)\left(\frac{\p x^{\bar{d}}}{\p x^b}B^b\right) = \frac{\p x^{\bar{c}}}{\p x^a}\frac{\p x^{\bar{d}}}{\p x^b}A^a B^b = \frac{\p x^{\bar{c}}}{\p x^a}\frac{\p x^{\bar{d}}}{\p x^b}T^{ab}.
\end{align}
This is the definition of a second order contravariant tensor. 

\section{Tensor Equations}
We can write a basic tensor equation,
\begin{align}
    T^a = k(A^a+B^a),
\end{align}
and wonder how this would look in a transformed coordinate system?
\begin{align}
    T^{\bar{b}} &= \frac{\p x^{\bar{b}}}{\p x^a}T^a = k\left(\frac{\p x^{\bar{b}}}{\p x^a}A^a + \frac{\p x^{\bar{b}}}{\p x^a}B^a\right) \\
                &= k(A^{\bar{b}}+B^{\bar{b}}).
\end{align}
So if a tensor equation is true, it is true in all coordinate systems.

\section{The metric tensor}
What is the metric? \emph{The metric is a measure of space.}
We define the metric tensor, 
\begin{align}
    g_{ab} &= \ve_a\cdot\ve_b = g_{ba},
\end{align}
so it is symmetric.
We can use this when calculating spacetime distances:
\begin{align}
    ds^2 &= \vec{dr}\cdot\vec{dr} = \left(dx^a\ve_a\right)\cdot\left(dx^b\ve_b\right) \\
         &= (\ve_a\cdot\ve_b)dx^a\,dx^b = g_{ab}dx^a\,dx^b.
\end{align}
Is it a tensor?
\begin{align}
    g_{\bar{a}\bar{b}} &= (\ve_{\bar{a}}\cdot\ve_{\bar{b}}) = \left(\frac{\p x^c}{\p x^{\bar{a}}}\ve_c\right)\cdot\left(\frac{\p x^{d}}{\p x^{\bar{b}}}\ve_d\right) \\
                       &= \frac{\p x^c}{\p x^{\bar{a}}}\frac{\p x^d}{\p x^{\bar{b}}}(\ve_c\cdot\ve_d) = \frac{\p x^c}{\p x^{\bar{a}}}\frac{\p x^d}{\p x^{\bar{b}}}g_{cd},
\end{align}
so it transforms as a tensor; a second order covariant tensor. 

\section{Kronecker Delta}
We can write an arbitrary vector as 
\begin{align}
    \vec{A} = A^a\ve_a = A^{\bar{b}}\ve_{\bar{b}} &= \left(\frac{\p x^{\bar{b}}}{\p x^a}A^a\right)\left(\frac{\p x^{d}}{\p x^{\bar{b}}}\ve_d\right) \\
                                                  &= \left(\frac{\p x^{\bar{b}}}{\p x^a}\frac{\p x^{d}}{\p x^{\bar{b}}}\right)A^a\ve_d = \left(\frac{\p x^{d}}{\p x^a}\right)A^a\ve_d \\
                                                  &= \tensor{\delta}{_a^d}A^a\ve_d = A^d\ve_d = A^a\ve_a
\end{align}

\chapter{}

\chapter{}

\chapter{}
Asbolute Derivative:
\begin{align}
    \frac{D\lambda^a}{ds} &= \frac{d\lambda^a}{ds} + \tensor{\Gamma}{^a_{bc}}\lambda^b\frac{dx^c}{ds}
\end{align}
Covariant Derivative:
\begin{align}
    \tensor{\lambda}{^a_{;c}} &= \frac{\p\lambda^a}{\p x^c}+\tensor{\Gamma}{^a_{bc}}\lambda^b
\end{align}
Christoffel Symbols:
\begin{align}
    \tensor{\Gamma}{^c_{ab}}\ve_c &= \frac{\p\ve_a}{\p x^b},\quad \tensor{\Gamma}{^c_{ab}} = \tensor{\Gamma}{^c_{ba}}
\end{align}
Other stuff:
\begin{align}
    \frac{\p g_{ab}}{\p x^c} &= \tensor{\Gamma}{^d_{ac}}g_{bd} + \tensor{\Gamma}{^d_{bc}}g_{ad}\\
    \frac{\p g_{bc}}{\p x^a} &= \tensor{\Gamma}{^d_{ba}}g_{cd} + \tensor{\Gamma}{^d_{ca}}g_{bd}\\
    \frac{\p g_{ca}}{\p x^b} &= \tensor{\Gamma}{^d_{cd}}g_{ad} + \tensor{\Gamma}{^d_{ab}}g_{cd}\\
    2\tensor{\Gamma}{^d_{ac}}g_{bd} &= \frac{\p g_{ab}}{\p x^c} + \frac{\p g_{bc}}{\p x^a} - \frac{\p g_{ca}}{\p x^b} \\
    \tensor{\Gamma}{^f_{ac}} &= \frac12 g^{fb}\left(\frac{\p g_{bc}}{\p x^a} - \frac{\p g_{ca}}{\p x^b} + \frac{\p g_{ab}}{\p x^c}\right) \\
                             &= \frac12 g^{fb}\left(\p_a g_{bc} - \p_b g_{ca} + \p_c g_{ab}\right)
\end{align}
We multiplied lefthandside of (6.7) by $\tensor{\delta}{^f_d}$.

\begin{example}[2D flat space]
    $x^A = \{x,y\}$:
    \begin{align}
        g_{AB} &= \begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix} = \text{diag}(1,1) \\
        \tensor{\Gamma}{^A_{BC}} &= 0 
    \end{align}
    So we don't have to deal with these in Cartesian coordinates. What about polar coordinates? $x^A = \{r,\theta\}$:
    \begin{align}
        ds^2 &= dr^2 + r^2\,d\theta^2 \\
        g_{AB} &= \text{diag}(1,r^2) \\
        \tensor{\Gamma}{^A_{BC}} \neq 0 
    \end{align}
    So we can still get non-zero Christoffel symbols even for flat space, but it is still "boring" really. 
\end{example}

Let's consider something more interesting, i.e. curved. 
For 3D space, we have
\begin{align}
    ds^2 &= dr^2 + r^2\,d\theta^2 + r^2\sin^2\theta\,d\phi^2 
\end{align}
But we want to use just the surface of a sphere, so fixed $r=a$:
\begin{align}
    ds^2 &= a^2\,d\theta^2 + a^2\sin^2\theta\,d\phi^2 = g_{AB}dx^A\,dx^B \\
    g_{AB} &= \text{diag}(a^2,a^2\sin^2\theta)
\end{align}
We have $g_{AB}$, but we want $g^{AB}$. Recall
\begin{align}
    g^{AB}g_{BC} &= \tensor{\delta}{^A_C}.
\end{align}
So we have a set of 4 simultaneous equations:
\begin{align}
    g^{A1}g_{1C} + g^{A2}g_{2C} &= \tensor{\delta}{^A_C}.
\end{align}
For diagonal $g_{AB}$ \textbf{ONLY}:
\begin{align}
    g^{AB}g_{BA} &= g^{AA}g_{AA} = 1 \implies g^{AA} = \frac{1}{g_{AA}} \\
    g^{AB} &= \text{diag}\left(\frac{1}{a^2},\frac{1}{a^2\sin^2\theta}\right)
\end{align}
So now we want to calculate
\begin{align}
    \tensor{\Gamma}{^\theta_{\theta\theta}} &= \frac12 g^{\theta B} \left(\p_\theta g_{B\theta} - \p_B g_{\theta\theta} + \p_{\theta} g_{\theta B}\right), \quad g^{\theta B} = 0, B\neq\theta \\
                                            &= \frac12 \frac{1}{a^2}\left(\p_{\theta} g_{\theta\theta} - \p_\theta g_{\theta\theta} + \p_\theta g_{\theta\theta}\right) =  0 \\
    \tensor{\Gamma}{^\theta_{\phi\theta}} = \tensor{\Gamma}{^\theta_{\theta\phi}} &= \frac12 g^{\theta B}\left(\p_\theta g_{B\phi} - \p_B g_{\phi\theta} + \p_\phi g_{\theta B}\right) \\
                                          &= \frac12 g^{\theta\theta}\left(\p_\theta g_{\theta\phi} - \p_\theta g_{\phi\theta} + \p_\phi g_{\theta\theta}\right) = 0 \\
    \tensor{\Gamma}{^\theta_{\phi\phi}} &= -\sin\theta\cos\theta \\
    \tensor{\Gamma}{^\phi_{\theta\phi}} &= \cot\theta
\end{align}
The rest of the Christoffel symbols for this example are 0 (there are $2^3=8$ in total?).

\section{Geodesic Equations}
The velocity is a tensor,
\begin{align}
    \vec{v} &= v^\alpha\ve_{\alpha} = \frac{\p x^\alpha}{\p\tau}\ve_\alpha
\end{align}
If there's no force, then there's no change in the velocity vector doesn't change, but its components might change. 
No force means the absolute derivative of the components: 
\begin{align}
    \frac{Dv^\alpha}{d\tau} &= 0
\end{align}
By an affine parameter, we mean a linear function of path length $u=A+Bs$, such as the proper time $\tau$.
\begin{align}
    \frac{dv^\alpha}{d\tau} + \tensor{\Gamma}{^\alpha_{\beta\gamma}}v^\beta\frac{dx^\gamma}{d\tau} &= 0 \\
    \frac{d^2x^\alpha}{d\tau^2} + \tensor{\Gamma}{^\alpha_{\beta\gamma}}\frac{dx^\beta}{d\tau}\frac{dx^\gamma}{d\tau} &= 0 \\
    \ddot{x}^\alpha + \tensor{\Gamma}{^\alpha_{\beta\gamma}}\dot{x}^\beta\dot{x}^\gamma &= 0
\end{align}
Let's guess and make a solution for the sphere, $s=a\theta$, so we are just going around the circumference of the sphere at constant $\phi$.
For $\theta$:
\begin{align}
    \frac{d^2\theta}{ds^2} + \tensor{\Gamma}{^\theta_{BC}}\frac{dx^B}{ds}\frac{dx^c}{ds} = 0 + \tensor{\Gamma}{^\theta_{\phi\phi}}\frac{d\phi}{ds}\frac{d\phi}{ds} = 0
\end{align}
We get a big tick and a gold star!
For $\phi$:
\begin{align}
    \frac{d^2\phi}{ds^2} + \tensor{\Gamma}{^\phi_{BC}}\frac{dx^B}{ds}\frac{dx^C}{ds} = 0 + \tensor{\Gamma}{^\phi_{\theta\phi}}\frac{d\theta}{ds}\frac{d\phi}{ds} + \tensor{\Gamma}{^\phi_{\phi\theta}}\frac{d\phi}{ds}\frac{d\theta}{ds} = 0 
\end{align}
So it's a geodesic path! Yayyyyyy!














\end{document}
















