\documentclass[a4paper, 11pt, normalem]{report}

\usepackage{../../../LaTeX-Templates/Notes}
\usepackage{subfiles}

\titlecontents{chapter}% <section-type>
    [0pt]% <left>
    {}% <above-code>
    {Lecture \thecontentslabel\quad}% <numbered-entry-format>
    {}% <numberless-entry-format>
    {\dotfill\contentspage}% <filler-page-format>
\titleformat{\chapter}{\fontsize{15}{17}\bfseries\normalfont}{\textbf{Lecture \thechapter}}{1em}{}
\titleformat{\subsubsection}{\fontsize{10}{13}\bfseries\scshape}{\textbf{\thesubsubsection}}{1em}{}
\setcounter{tocdepth}{4}
% \setcounter{secnumdepth}{1}

\renewcommand{\arraystretch}{1.2}

\newcommand\R{\mathbb{R}}
\newcommand\N{\mathbb{N}}
\newcommand\C{\mathbb{C}}
\newcommand\Z{\mathbb{Z}}
\newcommand\Rl{\mathsf{Re}}
\newcommand\Iy{\mathsf{Im}}
\newcommand\p{\partial}
\newcommand\ifnt{\int_{-\infty}^{\infty}}
\newcommand\ofnt{\int_{0}^{\infty}}
\newcommand\ifsum{\sum_{-\infty}^{\infty}}
\newcommand\F{\mathcal{F}}
\newcommand\La{\mathcal{L}}
\newcommand\K{\mathcal{K}}
\newcommand\om{\omega}
\newcommand\veca{\vec{a}(\vec{r})}
\newcommand\veru{\vec{r}(u)}
\newcommand\vr{\vec{r}}
\newcommand\vs{\vec{S}}
\newcommand\va{\vec{a}}
\newcommand\hn{\hat{n}}
\newcommand\hr{\hat{r}}
\newcommand\he{\hat{e}}
\newcommand\dy{\frac{dy}{dx}}
\newcommand\df{\frac{df}{dx}}
\newcommand\dyy{\frac{d^2 y}{dx^2}}
\newcommand\dff{\frac{d^2 f}{dx^2}}
\newcommand\dz{\frac{df}{dz}}
\newcommand\lam{\lambda}
\newcommand\e{\epsilon}
\newcommand\del{\nabla}
\newcommand\grad{\vec{\nabla}}

\title{Mathematics Workshop \vspace{-20pt}}
\author{Professor Ruth Gregory}
\date{\vspace{-15pt}Michaelmas Term 2018 - Epiphany Term 2019}
\rhead{\hyperlink{page.1}{Contents}}

\begin{document}

\maketitle
\tableofcontents

\part{Complex Analysis}
\chapter{The Complex Plane}
\section{Basics}
Course will use Riley, Hobson and Bence Chapters 3, 24, 25

Recall $i = \sqrt{-1}$ allows us to extend our notion of numbers as we go from a $\R$eal to the $\C$omplex plane.
\begin{equation}
    \C = \{z = x + iy\;|\;x,y \in \R\}
\end{equation}
We represent the complex plane with an argand diagram.
\begin{align}
    x &= \Rl(z) = \frac{z + \bar{z}}{2} \\
    y &= \Iy(z) = \frac{z - \bar{z}}{2} \\
    r &= \mathrm{mod}(z) = \sqrt{x^2 + y^2} \\
    \theta &= \mathrm{Arg}(z) = \arctan\left(\frac{y}{x}\right)
\end{align}
Notice that $e^{i\theta} = e^{i\theta + 2i\pi}$, so $\theta$ is not uniquely defined, so choose a range for $\theta$, e.g. $\theta \in [0, 2\pi), \theta \in (-\pi,\pi]$.

\section{Complex Functions}
\begin{itemize}
    \item $\C \to \C$
    \item $z \to f(z) = u(x,y) + iv(x,y)$
    \item Like two functions of real variables, e.g. $f(z) = z^2 = (x + iy)^2 = (x^2 - y^2) + 2ixy$
    \begin{itemize}
        \item $u(x,y) = x^2 - y^2$
        \item $v(x,y) = 2xy$
    \end{itemize}
\item Exponential: $e^z, \mathrm{exp}(z) = \sum_{n=0}^{\infty} \frac{z^n}{n!}$
\end{itemize}

\section{Branch Cuts}
For real numbers, $1^2 = (-1)^2 = 1$, you have two roots; but functions should be single valued. \\
$\sqrt{}$ is fine on the Real line - choose a root and stick to it, but for the Complex plane, say $\sqrt{1} = 1$, the disc around $z = 1$ means a loss of continuity. 
\begin{align}
    z &= re^{i\theta} \\
    \sqrt{z} &= \sqrt{r}e^{i\theta/2}, \theta \in (0, 2\pi)
\end{align}
We can choose this branch, but must cut the complex plane along the reals so that $\theta$ can't run higher than $2\pi$, or could choose
\begin{equation}
    \sqrt{z} = \sqrt{r}e^{i\theta/2 +i\pi} 
\end{equation}
but still need the same cut, so no ambiguity in definition of the function's square root. 
Here, the function's square root is double valued.
\begin{equation}
    \log(re^{i\theta}) = \log(r) + i\theta
\end{equation}
$\theta$ is not unique, $\theta + 2n\pi$ is also a legitimate answer. 
Again, cut the plane somewhere, decide on the branch of the log. 
Can cut on the positive reals or negatives. 
Branch choices appear around zeroes because polar coordinates are singular there: $\theta$ is not specified at $r = 0$.
We say $z=0$ is a branch point, indicated with a wavy line. 
Branch cuts from branch point either to infinity or another branch point. 

\begin{example}
\begin{equation}
    f(z) = \sqrt{z^4 + 1}
\end{equation}
Think of this through two steps: $z \to z^4 + 1 \to \sqrt{z^4 + 1}$. 
There is a branch cut around $z^4 + 1 = 0$. 
There are four branch points:
\begin{align}
    z^4 &= -1 \\
    z &= e^{i\pi/4}...
\end{align}
Around a branch point:
\begin{align}
    z &= e^{i\pi/4} + \epsilon e^{i\theta + 3i\pi/4} \\
    z^4 &= -1 + 4\epsilon e^{i\theta +3i\pi/4} \\
    \sqrt{z^4 + 1} &= 2\sqrt{\epsilon}e^{i\theta/2 + 3i\pi/8}
\end{align}
This gives rise to the same problem as circling the origin for $\sqrt{z}$.
This example has other choices for branch cuts, some may be able to limit ambiguities to limits. 
\end{example}

\section{Trig and Hyperbolic Functions}
Trig functions can be generalised to include complex numbers, and can be expressed as exponentials in the usual way, similarly with hyperbolics. 
$\sinh$ and $\cosh$ have periodicity of $2\pi i$.
\begin{align}
    \cos(z) &= \cosh(iz) \\
    \sin(z) &= i\sinh(iz)
\end{align}

\chapter{Complex Differentiation and Cauchy-Riemann}
\section{Continuity}
Definition:
\begin{equation}
    \lim_{z\to z_0} f(z) = w \in \C \iff \forall \e > 0 \;\exists \;\delta > 0 \; s.t.\; |z-z_0| < \delta \implies |f(z) - w| < \e
\end{equation}
\emph{If you are close to a point $z_0$, then f(z) is close to $f(z_0)$.}
f is continuous at $z_0 \iff \lim_{z\to z_0} f(z) = f(z_0)$.

Note that the limit must be path independent, and the real and imaginary parts must be continuous. 

\section{Differentation}
Recall
\begin{equation}
    \df = \lim_{h\to 0} \frac{f(x+h) - f(x)}{h}
\end{equation}
For $\C$ having two $\R$ dimensions, we have $\grad, \grad\cdot, \grad\times$. 
\begin{itemize}
    \item[X] Grad - acts on scalars ($\R$)
    \item[X] Div - acts on vectors ($\C$), gives scalar
    \item[X] Curl - in 3D
\end{itemize}
Try 
\begin{equation}
    \frac{df}{dz} = \lim_{\delta\to 0} \frac{f(z+\delta) - f(z)}{\delta}, \delta, z \in \C
\end{equation}
Require limit independent of direction.

\begin{example}
$f(z) = z^2$:
\begin{align}
    & \lim_{\delta\to 0} \frac{(z+\delta)^2 - z^2}{\delta} \\
    \implies & \lim_{\delta\to 0} \frac{\cancel{z^2} + 2z\cancel{\delta} + \delta^{\cancel{2}} - \cancel{z^2}}{\cancel{\delta}} = 2z
\end{align}
What about $f(z) = \bar{z}$?
\begin{equation}
    \frac{\bar{z} + \bar{\delta} - \bar{z}}{\delta} = \frac{\bar{\delta}}{\delta} = \exp{(-2i\mathrm{Arg}(\delta))}
\end{equation}
Not path independent, so no limit. 
\end{example}

$\frac{df}{dz}$ is the complex derivative of f at z, and f is differentiable at z if this limit exists. 

\section{Analytic Functions}
An analytic function is a complex function that is differentiable, at least in some region. 

Definition: A neighbourhood of $z \in \C$ is an open set U such that $z \in U$.

Definition: $f: \C \to \C$ is analytic/holomorphic at $z_0 \in \C$ if $\exists$ a neighbourhood U of $z_0$ on which f is differentiable $\forall \; z \in U$.

\begin{example}
    \begin{align}
        |z|^2 &= z\bar{z} \\
        \lim_{\delta\to 0} \frac{(z+\delta)(\bar{z} + \bar{\delta}) - z\bar{z}}{\delta} &= \bar{z} + \frac{\bar{\delta}}{\delta}z
    \end{align}
If $z \neq 0$, no limit, but $z = 0$ has limit, 0, $\implies$ differentiable at $z = 0$, \underline{but not analytic.}
\end{example}
$z^n$ is differentiable everywhere, so analytic on all $\C$.

\section{Cauchy-Riemann Equations}
We have
\begin{align}
    f(z) &= f(x + iy) \\
         &= u(x,y) + iv(x,y)
\end{align}
Derivation:
\begin{align}
    \frac{f(z_0 + \delta) - f(z_0)}{\underbrace{\delta}_{z - z_0}} &= \frac{u(x_0 + \delta x, y_0 + \delta y) - u(x_0,y_0)}{\delta x + i\delta y} + i\frac{v(x_0 + \delta x, y_0 + \delta y) - v(x_0,y_0)}{\delta x + i\delta y} \\
                                                                   &= \frac{u(x_0,y_0) + \delta x \frac{\p u}{\p x}|_0 + \delta y \frac{\p u}{\p y}|_0 - u_0}{\delta x + i\delta y} + i\frac{v_0 + \delta x \frac{\p v}{\p x}|_0 + \delta y\frac{\p v}{\p y} - v_0}{\delta x + i\delta y} \\
                                                                   &= \frac{\left(\frac{\p u}{\p x} + i\frac{\p v}{\p x}\right)\delta x + i\left(\frac{\p v}{\p y} - i\frac{\p u}{\p y}\right)\delta y}{\delta x + i\delta y}
\end{align}
For complex differentiation, limit must be independed of $\delta x + i\delta y$, so the numerator must factorise as $() \times (\delta x + i\delta y)$.
\begin{equation}
    \implies \frac{\p u}{\p x} + i\frac{\p v}{\p x} = \frac{\p v}{\p y} - i\frac{\p u}{\p y}
\end{equation}
This gives the C-R relations:
\begin{align}
    \frac{\p u}{\p x} &= \frac{\p v}{\p y} \\
    \frac{\p v}{\p x} &= -\frac{\p u}{\p y} 
\end{align}
Check with $z^2 = (x^2 - y^2) + 2ixy$:
\begin{align}
    \frac{\p u}{\p x} &= 2x & \frac{\p v}{\p y} &= 2x \\
    \frac{\p u}{\p y} &= -2y & \frac{\p v}{\p x} &= 2y 
\end{align}
Note:
\begin{equation}
    \frac{\p u}{\p x} = \frac{\p v}{\p y} \implies \frac{\p^2 u}{\p x^2} = \frac{\p^2 v}{\p x \p y} = -\frac{\p^2 u}{\p y^2}
\end{equation}
Or 
\begin{equation}
    \frac{\p^2 u}{\p x^2} + \frac{\p^2 u}{\p y^2} = \del^2 u = 0
\end{equation}
Suppose we regard z and $\bar{z}$ as independed variables. 
\begin{equation}
    x = \frac{z + \bar{z}}{2},~ y = \frac{z - \bar{z}}{2i} 
\end{equation}
\begin{align}
    \frac{\p}{\p \bar{z}} &= \frac{\p x}{\p \bar{z}}\frac{\p}{\p \p x} + \frac{\p y}{\p \bar{z}}\frac{\p}{\p y} = \frac{1}{2}\left(\frac{\p}{\p x} + i\frac{\p}{\p y}\right) \\
    \frac{\p}{\p z} &= \frac{\p x}{\p z}\frac{\p}{\p x} + \frac{\p y}{\p z}\frac{\p}{\p y} = \frac{1}{2}\left(\frac{\p}{\p x} - i\frac{\p}{\p y}\right) \\
    \frac{\p f}{\p \bar{z}} &= \frac{1}{2}\left(\frac{\p u}{\p x} + i\frac{\p v}{\p x} + i\frac{\p u}{\p y} - \frac{\p v}{\p y}\right) \\
                    &= \frac{1}{2}\left(\left(\frac{\p u}{\p x} - \frac{\p v}{\p y}\right) + i\left(\frac{\p v}{\p x} + \frac{\p u}{\p y}\right) \right) \\
                    &= 0 \iff \text{CR satisfied}
\end{align}
Can express analytically/differentiable as $\frac{df}{d\bar{z}} = 0$.

\chapter{Complex Integration and Cauchy's Theorem}
\section{Subsets of the Plane - Curves and Domains}
A continuous curve, $\gamma$, in $\C$ is a map.
\begin{equation}
    \gamma:~ [a,b] \subset \R \to \C
\end{equation}
$[a,b]$ is an interval, $\gamma(a)$ is the starting point, and $\gamma(b)$ is the final point. 
$-\gamma$ is the 'opposite' curve - it is the same path but opposite direction. 
\begin{equation}
    \gamma:~ x \in [0,2\pi] \to \C,~ z = e^{ix}
\end{equation}
This is a unit circle anti-clockwise. 
$e^{-ix}$ is $-\gamma$, a unit circle but clockwise. 

An open set G is said to be split if
\begin{equation}
    G = G_1 \cup G_2
\end{equation}
such that $G_1$, $G_2$ are open, not empty, and 
\begin{equation}
    G_1 \cap G_2 = \phi
\end{equation}
G is connected if it does not split. 
This means that any two points in G can be connected by a curve lying in G. 
\begin{equation}
    z_1,z_2 \in G: \exists\; \gamma \; s.t. \; \gamma_i = z_1, \gamma_f = z_2, \gamma \subset G.
\end{equation}
G is simply connected if any pair of curves connecting any pair of points in G can be continuously deformed into each other without leacing G.

\section{Complex Integration}
For real integration, there was no ambiguity in the path. For complex integration however, we have many paths. 

We must specify the curve of integration,
\begin{align}
    \int_\gamma f(z)\,dz &= \int_\gamma (u+iv)(dx + idy) \\
                         &= \int_\gamma (u\,dx - v\,dy) + i(v\,dx + u\,dy)
\end{align}
$dx$ and $dy$ are determined y $\gamma$.
\begin{align}
    dx &= x'(t)\,dt \\
    dy &= y'(t)\,dt 
\end{align}
$t \in [a,b]$, parameterising $\gamma$.

\begin{example}
    Find $\int_\gamma z^2\,dz$ for $\gamma: [0,\pi] \to \C, t \to e^{it}$.
    \begin{align}
        x &= \cos t \\
        y &= \sin t \\
        u &= \cos^2t - \sin^2t = \cos 2t \\
        v &= 2\sin t\cos t = \sin 2t \\
        \implies dx &= -\sin t\, dt \\
        \implies dy &= \cos t\,dt \\
        \int_\gamma u\,dx - v\,dy &= \int_{0}^{\pi} [-\cos 2t \sin t - \sin 2t\cos t]\,dt \\
                                  &= -\int_{0}^{\pi} \sin 3t\,dt \\
                                  &= \left[\frac{1}{3}\cos 3t\right]_{0}^{\pi} \\
                                  &= -\frac{2}{3} \\
        \int_\gamma v\,dx + u\,dy &= \int_{0}^{\pi} [-\sin 2t\sin t + \cos 2t\cos 2]\,dt \\
                                  &= \int_{0}^{\pi} \cos 3t\,dt \\
                                  &= 0
    \end{align}
    Note: $\int_{-1}^{1} x^2\,dx = [\frac{1}{3}x^3]_{-1}^{1} = -\frac{2}{3}$. 
    Also, $-\frac{2}{3}$ if $\gamma$ is lower semi-circle - the result seems to be path independent.
\end{example}

\section{Cauchy's Theorem}
Let $D \subset \C$ be a simply connected open subset of $\C$, and if $f: D \to \C$ is analytic on D, then for any closed curve $C \subset D$:
\begin{equation}
    \oint_C f(z)\,dz = 0.
\end{equation}
C indicates a closed curve. $\oint$ indicates integral round closed curve.
$\oint$ is a contour integral and C the contour of integration. 
Justify with real analysis. 

Recall Green's theorem:
\begin{equation}
    \oint_C \left[A\,dx + B\,dy\right] = \iint_S \left[\frac{\p B}{\p x} - \frac{\p A}{\p y}\right]\,dx\,dy
\end{equation}
Then 
\begin{align}
    \oint_C f(z)\,dz &= \oint_C [u\,dx - v\,dy] + i[u\,dy + v\,dx] \\
                     &= \iint_S \left[-\frac{\p v}{\p x} - \frac{\p u}{\p y}\right] - i\left[\frac{\p v}{\p y} - \frac{\p u}{\p x}\right].
\end{align}
Here, we use simple-connectedness.
But f is analytic in D, so analytic in S, hence by CR relations, this integral is 0. 
Note, if $\gamma_1,\gamma_2$ are 2 curves from $z_1 \to z_2$, $\gamma_1 - \gamma_2$ is a closed curve, hence
\begin{equation}
    \int_{\gamma_1} f = \int_{\gamma_2} f, \text{ for analytic f.}
\end{equation}
$C = \gamma_2 - \gamma_1$ closed. 

\emph{The integral of analytic functions are independent of the path.}

This also allows us to define an integral function as we now know it is path independent:
\begin{equation}
    F(z_2) - F(z_1) = \int_\gamma f\,dz,~ \frac{dF}{dz} = f.
\end{equation}
Finally, if $C_1$ and $C_2$ are closed contours enclosing surfaces $D_1$ and $D_2$, then if f is analytic on $D = D_2 - D_1$,
\begin{equation}
    \oint_{C_1} f(z_1)\,dz = \oint_{C_2}f(z)\,dz.
\end{equation}
Note that f doesn't need to be analytic on $D_1$ or $D_2$, e.g. $C_1 = e^{it}, C_2 = 2e^{it}, t \in [0,2\pi], f(z) = \frac{1}{z^2}$. This is not analytic on $D_1$ or $D_2$, but it is on $D_2 - D_1$.

\end{document}
